---
title: "Assignment 1"
author: "Shawn Ganz"
format: html
---

## Introduction

For this assignment I chose to use a csv from NYC OpenData called [2018 Central Park Squirrel Census - Squirrel Data](https://data.cityofnewyork.us/Environment/2018-Central-Park-Squirrel-Census-Squirrel-Data/vfnx-vebw/about_data) provided by "The Squirrel Census".

## Approach

Since the assignment is to transform this dataframe, I want to drop a couple of columns. Below is a list of all the columns:

```{r}
#| echo: false
df <- read.csv("https://raw.githubusercontent.com/Siganz/data_607_week_1/refs/heads/main/data/2018_Central_Park_Squirrel_Census_Squirrel_Data_20260126.csv")
colnames(df)
```

I want to create a dataframe with only these columns:

```{r}
#| echo: false
df <- read.csv("https://raw.githubusercontent.com/Siganz/data_607_week_1/refs/heads/main/data/2018_Central_Park_Squirrel_Census_Squirrel_Data_20260126.csv")
colnames(df)[c(3,9:11,16:20,1,2,31)]
```

Afterwards I want to create the following:

-   A binary “Active” squirrel column using the “Running,” “Chasing,” “Climbing,” “Eating,” and “Foraging” columns.

-   Convert the “Above Ground Sighter Measurement” column to numeric (INT/FLOAT) values only.

The motivation to use this dataset is simple, I just chose the first interesting popular dataset I found on NYC OpenData. This encourages an exploratory approach, which might be useful when learning new skills.

## Code-base

TODO: Code-base

## Video

TODO: Video
