---
title: "Assignment 1 – Loading Data into a Data Frame "
author: "Shawn Ganz"
format: html
---

## Introduction

For this assignment I chose to use a csv from NYC OpenData called [2018 Central Park Squirrel Census - Squirrel Data](https://data.cityofnewyork.us/Environment/2018-Central-Park-Squirrel-Census-Squirrel-Data/vfnx-vebw/about_data) provided by "The Squirrel Census".

## Approach

Since the assignment is to transform this dataframe, I want to drop a couple of columns. Below is a list of all the columns:

```{r}
df <- read.csv("https://raw.githubusercontent.com/Siganz/data_607_week_1/refs/heads/main/data/2018_Central_Park_Squirrel_Census_Squirrel_Data_20260126.csv")
colnames(df)
```

I want to create a dataframe with only these columns:

```{r}
colnames(df)[c(3,9:11,16:20,1,2,31)]
```

Afterwards I want to create the following:

- A binary “Active” squirrel column using the “Running,” “Chasing,” “Climbing,” “Eating,” and “Foraging” columns.

- Convert the “Above Ground Sighter Measurement”, "x", "y" columns to numeric (INT/FLOAT) values only.

The motivation to use this dataset is simple, I just chose the first interesting popular dataset I found on NYC OpenData. This encourages an exploratory approach, which might be useful when learning new skills.

## Code-base
### Setup
- read the data
- check data types

```{r}

library(tidyverse, ggplot2)

# Original Data
url <- "https://raw.githubusercontent.com/Siganz/data_607_week_1/refs/heads/main/data/2018_Central_Park_Squirrel_Census_Squirrel_Data_20260126.csv"

# Read Data
df <- read.csv(url, stringsAsFactors = FALSE)

# Optional, view first row (there's a lot of fields so it looks)
df[1,]

# Optional, view the field types
str(df)
```

Interesting that instead of using int/bool they used character. I would like to change that, while also creating vectors for the columns with different names. 

```{r}
# cols vector for field selection
# removed Lat.Long because you can get point file from x/y alone.

cols <- c(
  "Unique.Squirrel.ID",
  "Primary.Fur.Color",
  "Highlight.Fur.Color",
  "Combination.of.Primary.and.Highlight.Color",
  "Running",
  "Chasing",
  "Climbing",
  "Eating",
  "Foraging",
  "Above.Ground.Sighter.Measurement",
  "X",
  "Y"
)

# Copy
df2 <- df[ , cols]

# Check matrix
str(df2)
df2[1:5,]

# Change column names
colnames(df2) <- c(
  "unique_id",
  "primary_color",
  "highlight_color",
  "combination_color",
  "running",
  "chasing",
  "climbing",
  "eating",
  "foraging",
  "above_ground_measurement",
  "x",
  "y"
)

cols <- c(
  "unique_id",
  "primary_color",
  "highlight_color",
  "combination_color",
  "running",
  "chasing",
  "climbing",
  "eating",
  "foraging",
  "x",
  "y"
)

bool_cols <- c(
  "running",
  "chasing",
  "climbing",
  "eating",
  "foraging"
)

color_cols <- c(
  "primary_color",
  "highlight_color",
  "combination_color"
)
```

### Data viewing & Manipulation
I would now like to view the data and check out the uniques for certain columns and see what I can do with them.

```{r}
# check names
str(df2)
# check unique
for (col in bool_cols) {
  print(col)
  print(unique(df2[[col]]))
}
```

It seems that most of the columns are already clean, there might be an issue with the lowercase false/true so I'm going to upper those values.

```{r}
# the columns are clean, will use toupper() on the values 
for (col in bool_cols) {
  df2[[col]] <- toupper(df2[[col]])
}

str(df2)
```

I am using a for loop, this type of iteration is similar in python so it makes it simpler for me to remember. Now I would like to convert bool_cols into logical data types, using another loop.

```{r}
for (col in bool_cols) {
  df2[[col]] <- as.logical(df2[[col]])
}

str(df2)
```

Now I will create a new column called activity.

```{r}

df2$activity <- as.logical(rowSums(df2[bool_cols]) > 0)
str(df2)
```

I don't like that activity column is at the tail, moving it:

```{r}
df2 <- df2[, c(
  "unique_id",
  "primary_color",
  "running", "chasing", "climbing", "eating", "foraging",
  "activity", "above_ground_measurement",
  "x", "y"
)]

str(df2)

```

I  don't like the highlight_color and combination_color columns. I am going to drop them, while also checking for the unique in the remaining color column, primary_color.

```{r}
df2$highlight_color <- NULL
df2$combination_color <- NULL

# should return three lines, last two should just be NULL
for (col in color_cols){
  print(unique(df2[[col]]))
}

```

Now, I want to remove the empty string ("") from being a unique value. 

```{r}
df2$primary_color[df2$primary_color == ""] <- NA

unique(df2$primary_color)
```

Checking on uniques for the above_ground_measurement
```{r}
unique(df2$above_ground_measurement)
```

Looks like int, except for FALSE and "". So, we will look for any values %in% those and change them to "0" before converting to numeric. If this was a pipeline, I would build a function to normalize then find any non-numeric values.

```{r}
df2$above_ground_measurement[df2$above_ground_measurement %in% c("", "FALSE")] <- "0"

df2$above_ground_measurement <- as.numeric(df2$above_ground_measurement)

str(df2$above_ground_measurement)
```

Checking on X/Y to see if they have any non numeric symbols: 
```{r}
bad_rows <- which(grepl("^$|[^0-9.-]", df2$x))
bad_rows
bad_rows <- which(grepl("^$|[^0-9.-]", df2$y))
bad_rows
```

They aren't, so let's just do a simple as.numeric conversion

```{r}
df2$x <- as.numeric(df2$x)
df2$y <- as.numeric(df2$y)
str(df2)
```

This data looks good to me!

### Reporting

Now I would like to plot these using ggplot2, I had an AI generate the following code since I'm not familiar.


**Squirrel Behaviors**
```{r}
rates <- colSums(df2[bool_cols]) / nrow(df2)

plot_df <- data.frame(
  behavior = names(rates),
  proportion = as.numeric(rates)
)

ggplot(plot_df, aes(x = behavior, y = proportion)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Observed Squirrel Behaviors",
    x = "Behavior",
    y = "Percent of Observations"
  ) +
  theme_minimal()

```

**Share of Total Activity by Fur Color**
```{r}

counts <- aggregate(
  activity ~ primary_color,
  data = df2,
  FUN = sum,
  na.action = na.omit
)

counts$percent <- counts$activity / sum(counts$activity)

ggplot(counts, aes(x = primary_color, y = percent, fill = primary_color)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Share of Total Activity by Fur Color",
    x = "Primary Fur Color",
    y = "Percent of Total Activity"
  ) +
  theme_minimal() +
  theme(legend.position = "none")


```

*Summary Table for All Behaviors*
```{r}
# Create a summary table for all behaviors
do.call(rbind, lapply(bool_cols, function(col) {
  df2 %>%
    group_by(primary_color) %>%
    summarise(
      behavior = col,
      n_false = sum(get(col) == FALSE, na.rm = TRUE),
      n_true = sum(get(col) == TRUE, na.rm = TRUE),
      total = n(),
      .groups = 'drop'
    )
}))
```

**Behavior Rates by Primary Fur Color**
```{r}
# Calculate activity rates for all behaviors
activity_rates <- do.call(rbind, lapply(bool_cols, function(col) {
  df2 %>%
    filter(!is.na(primary_color)) %>%  # Remove NA colors
    group_by(primary_color) %>%
    summarise(
      behavior = col,
      n_true = sum(get(col) == TRUE, na.rm = TRUE),
      total = n(),
      activity_rate = n_true / total,
      .groups = 'drop'
    )
}))

# Create the bar chart
ggplot(activity_rates, aes(x = primary_color, y = activity_rate, fill = primary_color)) +
  geom_col() +
  facet_wrap(~ behavior, ncol = 3) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Behavior Rates by Primary Fur Color",
    x = "Primary Fur Color",
    y = "Percent Observed"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Coming from Python, I like f-strings, so here are some ones I made in R: 

```{r}
#| echo: false
paste(
  "Squirrel climbing stats:",
  colSums(df2["climbing"]),
  " / ",
  
  nrow(df2)
)

paste(
  "Squirrel eating stats",
  colSums(df2["eating"]),
  " / ",
  
  nrow(df2)
)

paste(
  "Squirrel foraging stats",
  colSums(df2["foraging"]),
  " / ",
  
  nrow(df2)
)

paste(
  "Squirrel running stats",
  colSums(df2["running"]),
  " / ",
  
  nrow(df2)
)

```


**Average Activity Rate by Primary Fur Color**
This one is pretty funny, I actually spent a good 30 minutes trying to figure out if my data was wrong, like if FALSE was somehow being aggregated, but it just turns out all the squirrels were very active!

```{r}
rates <- aggregate(
  activity ~ primary_color,
  data = df2,
  FUN = mean,
  na.action = na.omit
)

ggplot(rates, aes(x = primary_color, y = activity, fill = primary_color)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Average Activity Rate by Primary Fur Color",
    x = "Primary Fur Color",
    y = "Percent of Observations with Activity"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

### Code Base Conclusion
I went a little overboard, but I found it really helpful to learn more about r. I enjoy using the qmd since the "{r}" is easy to type and I can always check the output via html or using the console in rstudio. There is obviously a lot that I need to work on (ggplot2, aggregate, tidyverse), however this was a great introduction. I actually enjoy that base r is very intuitive and I feel like I can pick this up rather quickly. 

## Video

TODO: Video
